A Simple Type Theory for Category Theory

### 2022-05-25 paper/paper.tex
some "category theories"

Since many type theories are given semantics by free structured
categories, we can use SCT as a logical framework in which to do
``formal type theory''.

>>>>

Since many type theories are given semantics by free structured
categories, we can use SCT as a logical framework in which to do
``formal type theory'' by axiomatizing the structure of the categories
involved. The theories we can define are limited because we do not
have access to opposite categories, but we can define some modal and
monoidal structures.

\begin{definition}
  An \emph{SCT-signature} is a well-formed context $\Gamma \isadtctx$.
\end{definition}

\begin{lemma}
  Any \emph{unary mode theory} can be defined as an SCT-signature as follows.
  \begin{enumerate}
  \item For each mode $m$ add an assumption $m : \Cat$
  \item For each 1-morphism $f : m \to n$ add an assumption $R_f :
    \harr m n$ and assumptions $F_f : \varr m n$ with and $G_f : \varr
    n m$ with assumptions that they represent $R_f$.
  \item For each 2-morphism $\alpha : f \Rightarrow g$ add an assumption $\alpha : R_f \multimap R_g$
  \item Add each equality as an assumption using the identity type Id.
  \end{enumerate}
\end{lemma}

This translation illustrates some ways the notion of SCT-signature is
more flexible than a mode theory. For instance, in an SCT-signature we
can add a profunctor assumption $R_f : \harr m n$ without saying
whether it is representable as $F_f$ or $G_f$, and we can add more
complex 2-morphisms.

\begin{definition}
  The \emph{theory of a strict monoidal category} consists of
  \begin{enumerate}
  \item A category $C : \Cat$
  \item A unit $I : \varr 1 C$
  \item A tensor product $\otimes : \varr {(C \times C)} C$
  \item Such that $\alpha:(C \times C) \times C \vdash (\pi_2\pi_1\alpha \otimes \pi_2\pi_1 \alpha) \otimes \pi_2 \alpha = \pi_2\pi_1\alpha \otimes (\pi_2\pi_1 \alpha \otimes \pi_2 \alpha)$
    and $\alpha \otimes I() = \alpha = I() \otimes \alpha$
  \end{enumerate}
\end{definition}

\begin{definition}
  The theory of a cartesian strict monoidal category is an extension
  of the theory of a strict monoidal category with representability
  data
  \[ \punitinXfromYtoZ {C} {\Gamma}{\Delta \otimes \Sigma} \cong (\punitinXfromYtoZ C \Gamma \Delta)\times (\punitinXfromYtoZ C \Gamma \Sigma) \]
  and
  \[ \punitinXfromYtoZ {C} {\Gamma} I \cong 1 \]
\end{definition}

\begin{definition}
  The theory of a simple type theory is an extension of the theory of
  a strict cartesian monoidal category with a
  \begin{enumerate}
  \item A category $\text{Ty}$ of types
  \item A profunctor $\vdash : \harr C {\textrm{Ty}}$
  \item A functor $\text{var} : \varr{\textrm{Ty}} C$ representing $\vdash$ :
    \[ \Gamma \vdash A \cong \punitinXfromYtoZ {} \Gamma {\text{var} A} \]
  \item As assumption that $\text{var}$ is fully faithful:
     \[ \punitinXfromYtoZ {} {\text{var} A} {\text{var} B} \cong \punitinXfromYtoZ {} A B \]
  \end{enumerate}
\end{definition}

\begin{definition}
  The theory of CBPV is an extension of the theory of a simple type
  theory $(\text{VCtx},\times,1,\text{VTy},\vdash_V,\text{var}_V)$ with
  \begin{enumerate}
  \item A category $\text{CCtx}$
  \item An object $I : \varr 1 {\text{CCtx}}$
  \item An action of $\text{VCtx}$ on $\text{CCtx}$, i.e.,
    \begin{enumerate}
    \item A functor $\oslash : \text{VCtx}\times \text{CCtx} \to \text{CCtx}$
    \item Such that $1 \oslash \underline\Delta = \underline\Delta$
    \item And $(\Gamma \times \Delta) \oslash \underline\Sigma = \Gamma \oslash \Delta \oslash \underline\Sigma$
    \end{enumerate}
  \item A category $\text{CTy}$ with a profunctor $\vdash_C : \harr
    {\text{CCtx}}{\text{CTy}}$ and a fully faithful functor $\text{var}_C$ that represents it
    \[ \underline\Delta \vdash_C \underline B \cong \punitinXfromYtoZ {} {\underline\Delta}{\text{var}_C\underline B}\]
  \end{enumerate}
  Furthermore we can add in the universal properties of any
  non-contravariant CBPV connectives. For instance:
  \begin{enumerate}
  \item $F : \varr {VTy} {CTy}$ satisfying
    \[ \Gamma \oslash \text{var}_C F A \vdash_C \underline B \cong \Gamma \times \text{var}_V A \oslash I \vdash_C \underline B  \]
    \[ \Gamma \vdash_V U \underline B \cong \Gamma \oslash I \vdash_C \underline B  \]
  \end{enumerate}
\end{definition}
